---
title: "joint spp distance sampling"
author: "jmm"
date: "5/9/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The goal here is to model changes in species abundances based on distance sampling. To model species responses, we want to use a joint species modelling approach a la Ovaskainen.

As a starting point we need a distance sampling model that can be fitted with `Stan` so that we can then include the joint species part.

Distance sampling is a good example of a hierarchical model where we separate the ecological quantities of interest (animal abundance/density), from the observation of such quantities.

Let's start by simulating a data set and trying to recover known parameters. We assume that detection probability decreases with distance to the trasect as a half-normal: $\exp\left(- \frac{x^2}{2\sigma^2} \right)$

```{r, echo=FALSE, fig.width=3, fig.height=3}
curve(exp(-x^2/1), xlim=c(0,3), xlab="distance", ylab="detection prob", 
      xaxt='n', lwd=2, las=1, bty = "l")
```

We assume further that a total of $n$ individuals are located at random over the area being sampled. We run the transect for a length $L$ and the boundaries of the study area are at a distance $B$ from each side of the transect. Let's simulate some data:

```{r}
set.seed(36)
sigma = 1
n = 200
B = 3
L = 10

x <- runif(n, -B, B)
p <- exp(-x^2/2*sigma^2) # detection probability
y_obs <- rbinom(n, size = 1, prob = p)
```

Now we plot the distribution of distances of all the animals to the transect and the distance of those that were obseved:

```{r,  fig.width=4, fig.height=3}
hist(abs(x), nclass = 15,
     xlab = "Distance (x)", col = "grey", 
     main = "True (grey) \nand observed distances (blue)" )
hist(abs(x[y_obs==1]), col = "blue", add = TRUE)
```

To estimate density $\frac{n}{ L \times 2 B}$ we can use a data augmentation approach as presented in [Royle and Dorazio (2009)](https://books.google.co.uk/books?id=rDppWpVP6a0C&lpg=PA236&ots=uWjd9RbkNV&dq=WinBUGS%20specification%20of%20distance%20sampling%20model%20for%20the%20impala%20data%2C%20with%20the%20half-normal%20detection%20function.&pg=PA237#v=onepage&q&f=false). For this, we add to the list of observed individual a number $nz$ of zeros and define a parameter $\psi$ for the probability that an individual (observed or not) is part of the population under study.

Because with `Stan` we cannot have discrete-valued hidden variables we have to use a zero-inflated version, see the [dist.stan](https://github.com/jmmorales/distance/raw/master/dist.stan) file

Let's fit this model to the simulated data

```{r, cache=TRUE}

nz = 200

datos <- list(n_obs = sum(y_obs),
              nz = nz,
              x = abs(x[y_obs==1]),
              y = c(rep(1,sum(y_obs)), rep(0,nz)),
              B = B,
              Area = 2*B*L)

pars = c("D", "psi", "sigma")

library("rstan")
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

fit <- stan(file = 'dist.stan',
            data = datos,
            pars = pars,
            include = TRUE,
            iter = 1000, thin = 1, chains = 3)

print(fit)

samples = extract(fit)

op = par(mfrow=c(1,2))
plot(density(samples$D), main = "density", xlab = "")
abline(v = n/(2*B*L))
plot(density(samples$sigma), main = "sigma", xlab = "")
abline(v=sigma)
par(op)
```

## Integrating the detection function

An alternative way of estimating $\sigma$ and population density is by looking at the probabilities of obtaining the observed distances $x$. Using Bayes rule, we can see that 

$$
p(x|y=1) = \frac{p(y=1|x)p(x)}{p(y=1)}
$$

for the case of the half-normal, we get
$$
p(x|y=1) = \frac{\exp(- x^2/(2 \sigma^2))}{\int_0^B{\exp(- x^2/(2 \sigma^2))}}
$$

We can use the function `integrate_1d` in `Stan` to get the denominator. See the [dist_int.stan](https://github.com/jmmorales/distance/raw/master/dist_int.stan) file (this is a conditional likelihood model and not the full likelihood one). 

Let's fit this model to the simulated data

```{r, cache=TRUE}

datos <- list(n_obs = sum(y_obs),
              x = abs(x[y_obs==1]),
              B = B,
              Area = 2*B*L)

pars = c("D", "sigma", "pbar")

library("rstan")
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

fit <- stan(file = 'dist_int.stan',
            data = datos,
            pars = pars,
            include = TRUE,
            iter = 1000, thin = 1, chains = 3)

print(fit)

samples = extract(fit)

op = par(mfrow=c(1,2))
plot(density(samples$D), main = "density", xlab = "")
abline(v = n/(2*B*L))
plot(density(samples$sigma), main = "sigma", xlab = "")
abline(v=sigma)
par(op)
```

## Joint Species Model

Now let's look at a case where we have several species. Following [Ovaskainen et al. 2019](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.2622), the number of species is denoted by $n_s$, the number of species traits by $n_t$, and the number of species-specific parameters by $n_p$. The species-specific parameters are combined into the $n_s \times n_p$ matrix $\mathbf{\Theta}$, so that each row of $\mathbf{\Theta}$, i.e., the vector $\mathbf{\Theta}_s$, contains the parameters for species $s$. The number of parameters ($n_p$) and their interpretations depend on the specific model.

To determine the way in which species respond to environmental factors (as described by the parameters $\mathbf{\Theta}$) depend on their traits and phylogenetic relationships, we vectorize the matrix $\mathbf{\Theta}$ to the $n_s n_p \times 1$ vector $\mathbf{\theta} = vec(\mathbf{\Theta})$, and model it using a multivariate normal distribution as

$$
\theta \sim \text{N} \left(\mathbf{m}, \Sigma \otimes \left[ \rho \mathbf{C} + \left(1 - \rho \right) \mathbf{I}_{n_s} \right] \right)
$$

here the $n_s n_p \times 1$ vector $\mathbf{m} = \text{vec}(\mathbf{M})$ is the vectorized version of the $n_s \times n_p$ matrix $\mathbf{M}$ (with elements $m_{sp}$, where $p = 1,...,n_p$ is the index for the regression parameter), and it gives the expected  parameters based on species traits. The influence of traits is modeled with a linear regression $m_{sp} = \sum_k t_{sk} \zeta_{kp}$ ,where the index $k$ runs over then $t$ traits, $t_{sk}$ is the trait $k$ for species $s$, and the parameter $\zeta_{kp}$ measures the influence of trait $k$ on parameter $p$. Arranging the regression parameters $\zeta_{kp}$ into a $n_t \times n_p$ matrix $\mathbf{Z}$ and the trait values $t_{sk}$ into a $n_s \times n_t$ matrix $\mathbf{T}$, we can write in matrix form $\mathbf{M} = \mathbf{T} \mathbf{Z}$. To include the intercept intothe model, we set $t_{s1}=1$ for all species $s$.

The $n_p \times n_p$ variance–covariance matrix $\Sigma$ above models the species-specific deviations around the expectation based on species traits, and $\otimes$ is the Kronecker (outer) product. The $n_s \times n_s$ matrix $\mathbf{C}$ is a phylogenetic correlation matrix that can be derived from a phylogenetic tree based on genetic data, or constructed from taxonomic information if a quantitative phylogeny is not available. The matrix $\mathbf{I}_{n_s}$ is the identity matrix of dimension $n_s$, and the parameter $0 \leq \rho \leq 1$ measures the strength of the phylogenetic signal.

We can use this setup to model the number of individuals of species $s$ at sample $i$ as

$$
\begin{aligned}
& y_{is} \sim \text{Poisson}(\lambda_{is}) \\
& \log(\lambda_{is}) = \theta_{0s} + \theta_{1s} \times x_i \\
\end{aligned}
$$

where parameters $\theta_{0}$s and $\theta_{1}$s are sampled from a multivariate normal with mean related to species traits and variance-covariance related to the phylogenetic similarity among species as described above.

### Simulation

Lets simulate some data. For this, we start by defining the parameter structure as described above. We will have three parameters as we will use one for the scale of detection probability.

```{r}
library(ape)
library(mvtnorm)
set.seed(36)

n_s <- 20  # number of spp
n_p <- 3   # number of parameters
n_t <- 2   # number of traits

# define species traits
dgrass <- rbeta(n_s, 1, 1)  # some variable such as fraction of grass in diet
log_bm <- rnorm(n_s, 0, 1)  # log body mass

# define trait matrix includding ones for the intercept
TT <- as.matrix(cbind(rep(1, n_s), scale(dgrass), scale(log_bm)))

# simulate phylogeny 
tree <- rtree(n = n_s)    
CC <- vcv(tree, corr = TRUE) # species correlation based on phylogeny

# sort species and re-arrange phylogenetic correlation
tmp <- dimnames(CC)
ids <- as.numeric(as.factor(tmp[[1]]))

C <- matrix(NA, ncol(CC), ncol(CC))
for(i in 1:ncol(CC)){
  for(j in 1:ncol(CC)){
    C[ids[i],ids[j]] <- CC[i,j]
  }
}

# define parameters that link traits to expected thetas
Z <- matrix(c(1, 0, 0.2, 0.1, 0.5, 0.4, 0.5, -0.5, 0 ), (n_t + 1), n_p)

# get expected thetas
M <- TT %*% Z

Sigma <- diag(n_p) * 0.5
rho = 0.5  # correlation with phylogeny

thetas <- rmvnorm(1, mean = as.vector(M), kronecker(Sigma, rho*C + (1-rho) * diag(n_s)))
Theta <- matrix(thetas[1,], n_s, n_p)
```

We can inspect now how grass in diet and body mass relate to the intercept and slope of the regression between the predictor $x$ and the expected count at a samping unit
```{r}
nf = layout(matrix(c(1,2,3,4,0,0),3,2,byrow=TRUE), widths=c(1,1), heights=c(1,1,0.1))
#layout.show(nf)

op <- par( mar = c(3, 3, 2, 2) + 0.1, mgp = c(3.5, 1, 0), las = 1, bty = "n", cex = 1.2)
plot(scale(dgrass), Theta[,2], ylab = "", xlab = "", main = "intercept")
plot(scale(dgrass), Theta[,3], ylab = "", xlab = "", main = "slope")
plot(scale(log_bm), Theta[,2], ylab = "", xlab = "")
plot(scale(log_bm), Theta[,3], ylab = "", xlab = "")
mtext("         scaled grass in diet           scaled log body mass", side = 1, line = -2, outer = TRUE, cex=1.3)
par(op)
```


Now we define the sampling scheme. Here we follow the simulation sheme of package `AHMbook`.

```{r}
n_sites <- 100 # sampling units
# sample level predictor (environmental covariate)
x <- rnorm(n_sites)
B = 5 # max distance

# species indicator
j <- rep(1:n_s, each = n_sites) 
site <- rep(1:n_sites, n_s)
x.s <- rep(x, n_s)
n <- length(j) # sample size

log_lambda <- Theta[j,1] + Theta[j,2] * x.s
N <- rpois(n, exp(log_lambda))
sigma <- exp(Theta[,1])
```

```{r}
data <- NULL
for (i in 1:length(N)) {
  d <- runif(N[i], 0, B)
  p <- exp(-d * d/(2 * (sigma[j[i]]^2)))
  y <- rbinom(N[i], 1, p)
  d <- d[y == 1]
  y <- y[y == 1]
  
  if (sum(y) > 0){
    data <- rbind(data, cbind(rep(site[i], sum(y)), rep(j[i], sum(y)), y, d))
  } 
}
colnames(data) <- c("site","sp", "y", "d")
datos = as.data.frame(data)
```

Now call `Stan`

```{r}

library("rstan")
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

stan_dat <- list(
  n_obs = dim(datos)[1],
  n_sites = n_sites,
  B = B,
  site = as.integer(datos$site), # (c(s, rep(1:n_sites, each = nzs ))),
  r = datos$d,
  K = 2, 
  X = as.matrix(cbind(numeric(length(x))+1, x)),
  n_max = rep(100, n_s),
  n_s = as.integer(n_s),
  n_t = 3,
  TT = TT,
  C = C,
  ones = numeric(n_s) + 1,
  sp = datos$sp
)

pars <- c("pbar", "b_m", "rho",  "Sigma", "z")


fit <- stan(file = 'dist_pois_bin_int.stan',
            data = stan_dat,
            pars = pars,
            iter = 1000, thin = 1, chains = 3)

#print(fit)

```


Before making any inferences we check for convergence and reasonable sample sizes on the posteriors

```{r}
fit_summary <- summary(fit)$summary
op <- par(mfrow = c(1,2))
hist(fit_summary[,10], main = "R-hat")
hist(fit_summary[,9], main = "n-eff" )
par(op)

N <- dim(fit_summary)[[1]]
for (n in 1:N) {
  rhat <- fit_summary[,10][n]
  if (rhat > 1.1 || is.infinite(rhat) || is.nan(rhat)) {
      print(sprintf('Rhat for parameter %s is %s!',
                    rownames(fit_summary)[n], rhat))
    }
  }
```

Rhats and n-eff look good (we get a waringin for the Rhat of Omega[1,1] but that's a fixed value so no worries).

We can now check some posteriors such as that for $\rh0$

```{r}
draws <- extract(fit)

plot(density(draws$rho), main = "")
abline(v=rho)
```

Now compare true $\zeta$ values (in red) to fitted ones
```{r}
zs <- fit_summary[grepl("z", rownames(fit_summary)),]
#plot(c(Z) - zs[,1])
df <- data.frame(x = 1:dim(zs)[1],
                 tz = c(Z),
                 fz = zs[,1],
                 L = zs[,4],
                 U = zs[,8])

ggplot(df, aes(x = x, y = tz)) +
  geom_point(size = 2, color="red") +
  geom_point(aes(y = fz), size = 2) +
  geom_linerange(aes(ymin = L, ymax = U)) +
  theme_classic()

```


Let's compare the sample-level regression coefficients
```{r}
bs <- fit_summary[grepl("b_m", rownames(fit_summary)),]

nf = layout(matrix(c(1,2,3,4,0,0),3,2,byrow=TRUE), widths=c(1,1), heights=c(1,1,0.1))
#layout.show(nf)
op <- par( mar = c(3, 3, 2, 2) + 0.1, mgp = c(3.5, 1, 0), las = 1, bty = "n", cex = 1.2)
plot(scale(dgrass), Theta[,2], col = 2,  ylab = "", xlab = "", main = "intercept")
points(scale(dgrass), bs[seq(1, (n_s*3) ,by=3) + 1,1])

plot(scale(dgrass), Theta[,3], col = 2 , ylab = "", xlab = "", main = "slope")
points(scale(dgrass), bs[seq(1, (n_s*3) ,by=3) + 2,1])

plot(scale(log_bm),Theta[,2], col = 2 , ylab = "", xlab = "")
points(scale(log_bm), bs[seq(1, (n_s*3) ,by=3) + 1,1])

plot(scale(log_bm),Theta[,3], col = 2 , ylab = "", xlab = "")
points(scale(log_bm),  bs[seq(1, (n_s*3) ,by=3) + 2,1])
mtext("         scaled grass           scaled log body mass", side = 1, line = -2, outer = TRUE, cex=1.3)
par(op)
```



